---
title: "Amiga Disk File Explorer"
author: "Pepijn de Vries"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Amiga Disk File Explorer}
  %\VignettePackage{adfExplorer} 
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Although the adfExplorer package is well documented, information is
scattered over various methods, functions and data. This vignette
attempts to provide a more complete overview of the package and its
background.

## Amiga Disk Files - Introduction

The Commodore Amiga was a home computer that relied heavily on it's
build-in 3.5 inch floppy disk drive in the late eighties and early
nineties. The Amiga was a popular system which nowadays can 
be accurately [emulated](https://en.wikipedia.org/wiki/Amiga_emulation)
on modern machines. A problem with the emulation of the Amiga
is that most modern machines no longer have floppy disk drives.
And if they do, they are usually physically not able to read
Amiga formatted disks. To overcome this problem, the Amiga Disk File
(ADF) was created. Such files are a virtual representation of a
floppy disks which can be used in emulation.

### Legal issues

Emulating the Commodore Amiga hardware is, as far as I can tell, legal.
However, the Amiga had its operating system stored on a read only memory
chip. Which is commonly referred to as the kickstart. This kickstart software
is still protected by copyright law. Licensed copies can be bought legally
from [AmigaForever](https://www.amigaforever.com). The same is true for much of the
Amiga software, which is still protected by copyright.

Note that for analysing ADF files with the adfExplorer package you don't
need the kickstart software. Just make sure the ADF files you use contain
data or software for which you own (or don't need) a licence. 

## The adfExplorer package

This package will allow you to analyse Amiga Disk Files (ADF) with
the R scripting language. It can also be used to transfer files
from and to ADF in batches, e.g. for preparing disks for emulation 
purposes.

The focus will be on the older Amiga operating systems (i.e.,
OS 3.x and less), as I am most familiar with those systems. In
newer versions floppy disks became less important anyway.
Note that this package cannot read extended ADF files containing
information on the disk's
[Modified Frequency Modulation (MFM)](https://en.wikipedia.org/wiki/Modified_Frequency_Modulation).
This information is typically only required for copy protected disk's
and is therefore out of the scope of this package.

### Future developments

The package is still maturing. The foundation of the package is solid
and it provides the essential means to access ADF files, there is still
room for improvement. The following improvements are planned for future
releases of the package:

* check and optimise current code;
* include better validation routines;
* delete files or directories from a virtual disk;
* retrieve and or modify file and directory details;
* optimise disk structure (when information is fragmented).

The usefulness of this package will increase with better support of Amiga
file formats. I do have plans to develop a separate package for supporting
typical Amiga file formats. I'm not sure when the first release will be
available, as I work on these projects in my spare time. For now the
ProTracker module format is supported in R
(https://cran.r-project.org/package=ProTrackR).

## Technical background

Some technical details will be presented here, as they will help
you better understand some of the package functionalities. I will
not discuss the ADF format in depth. A more detailed description
can be found in the [FAQ](http://lclevy.free.fr/adflib/adf_info.html)
by Laurent Cl&eacute;vy. In fact, I used that document as the main
source of information during the development of this package; it is
a very accurate description of the ADF format.

A physical Amiga floppy disk is divided into 80 cylinders, on 2 tracks
or sides, which contain 11 (double density (DD) disk) or
22 (high density (HD) disk) sectors or blocks. So in total a disk is
composed of either 1,760 or 3,520 blocks. Each block holds 512 bytes
of information. An Amiga Disk File (ADF) is dump of such ordered
blocks in a file. The adfExplorer package represents ADF files as
`amigaDisk` objects. A completely blank disk can be created using the
object constructor (`new`).

```{r}
library(adfExplorer)
blank.disk <- new("amigaDisk")
```

The blocks on Amiga disks are represented by `amigaBlock` objects, which
can be created with the constructor, or it can be extracted from an `amigaDisk`
object.

```{r}
## Create with constructor:
blank.block <- new("amigaBlock")

## Extract the first block from an amigaDisk object:
blank.block <- amigaBlock(blank.disk, block = 0)
```

The relationship between the physical location on the disk and the block
identifier can be obtained with the `get.blockID` method.

```{r}
get.blockID(disktype = "DD", sector = 4, side = 0, cylinder = 35)
```

Note that the identifiers have a base index of zero as opposed to one used
by R. The physical location can also be calculated from the block
identifier with the `get.diskLocation` method.

```{r}
get.diskLocation(disktype = "DD", block = 1231)
```

The completely `blank.disk` created earlier with the constructor can hold any
kind of data. However, in order to hold files, the disk needs to be formatted
with a specific file system. That way, the Amiga Disk Operating System (DOS)
can access it. We can use the `blank.amigaDOSDisk` method to create a blank
disk, structured such that it can hold files.

```{r}
blank.disk <- blank.amigaDOSDisk(diskname = "empty")
print(blank.disk)
```

As you can see, although it is blank, it is not completely empty. This is
because the file system also consumes disk space. To understand this you
need to know a bit more about the Amiga's file system and the way information
is stored in blocks on the disk. This will be discussed to some extent.

### File systems on the Amiga {#filesystem}

With Amiga OS version 2.0, the so-called Fast File System (FFS) was introduced.
The previous file system was not named at the time, but is now commonly
referred to as the Old File System (OFS). The OFS sacrifices disk space
for validation purposes, making it more robust in case of data recovery when
a disk got damaged. This advantage was dropped with the FFS making it
slightly faster (on original machines) and gaining disk space for file
data. The FFS does not have backward compatability. So disks formatted with
this file systems cannot be read by Amiga OS versions <2.0.

In OS version 2.0 the 'international mode' was also introduced. This mode was
meant to correct for a mistake in the routine to convert text into upper case.
On the Amiga file names can have both lower and upper case characters. But
during file name matching, the case is ignored. For that purpose, file names
are shifted to upper case in file name matching routines.

The Amiga uses the ISO 8859 Latin-1 character set, where in older
operating systems (<2.0), international characters (e.g., '&oslash;')
were not capitalised. This mistake was corrected in OS 2.0, but is
optional. In combination with the 'directory cache mode' (see below),
the international mode is mandatory.

With Amiga OS 3.0, the '[directory cache mode](#dircacheblock)' was
introduced. With the direct cache mode, one or more blocks are stored
for each directory (including the root) with basic information about
the files stored in that directory. In older versions, the directory
[header](#headerblock) only stored pointers to the files in that directory.
This meant that in older OS versions the header of each file needed to be
loaded to list all files in that directory. As with the directory cache mode
all information was stored in one (or more when necessary) block, it
was faster at listing directory content. On the original machine that
is, as floppy disk drives were pretty slow.

Information is stored in separate blocks, where each block is composed of 512
bytes. So even when a file is 10 bytes long, it will still consume at least
512 bytes of disk space. There are several different types of blocks,
carnying different types of information. Some of them will be discussed below.

Instead of working with a blank disk, it can be more informative to work
with a disk with some data on it. From here on let's work with the example
`amigaDisk` object provided with the package.

```{r}
data("adf.example")
```

### The bootblock {#bootblock}

The bootblock are actually the first two blocks on an Amiga disk:
`amigaBlock(adf.example, 0)` and `amigaBlock(adf.example, 1)`. When the
Commodore Amiga system boots, it will first load these blocks in memory
and uses this to check what type of disk it is. Any executable code on this
block will be run when the bootblock checksum is valid. When the system is
already booted when the disk is inserted, the bootblock will just be used to
determine what type of disk it represents.

### The root block

The root block holds information on the files and directories in the root of
the disk. It also holds information on the disk name, creation and modification
date. It is usually situated at the centre of the blocks, at block 880 for
DD disks.

```{r eval=FALSE}
amigaBlock(adf.example, 880)
```

### The bitmap block

The bitmap block, contains information about which blocks on the disk are in use
and which are free. The first four bytes in this block form a checksum, the
following 220 bytes (for a DD disk, twice as many for a HD disk). Each byte
is composed of 8 bits. When a bit is set, it represents a used block, when it
is not set, it reflects a free block. The bitmap block is usually situate next
to the root block, at block 881 on a DD disk.

```{r eval=FALSE}
amigaBlock(adf.example, 881)
```

The bitmap is thus used to allocate free blocks on the disk to put new files.
When files are removed (and blocks) are thus freed, this will also be marked
in the bitmap. The bitmap is also used to calculate the free space on a disk.

### Header blocks {#headerblock}

As indicated before, files are stored as 512 blocks. But not only the file
data is written to the disk. Information about the file (i.e., the file name,
it's size, where to find the data on the disk etc.) also needs to be stored.
This block, containing information on the file (or directory) is called a
header block. Each file, directory or link on the disk starts with a header
block. The root block is a special case of a header block; it marks the root
directory.

Header blocks of directories contain pointers to the files (and directories)
in that directory. Header blocks of files contain pointers to data blocks,
containing the data of the file.

### Data blocks

In the Fast [File System](#filesystem), the data blocks are only composed of
file data; nothing else. The location of the data blocks are provided in the
file's header. In the Old [File System](#filesystem), each data block contains
additional information, most importantly: a pointer to the next data block and
a checksum.

### The directory cache block {#dircacheblock}

The directory cache block is a block listing the most important information
of files and directories within a specific directory. This block type was
introduced with the [direct cache mode](#filesystem) in Amiga OS 3.0. It
basically stores a summary of header block information of all files in a
specific directory at a central place, making directory listing faster.

There is no backward compatibility for the directory cache mode, meaning that
OS versions <3.0 are not able to read disks that are formatted in this mode.
The directory cache mode is always used in combination with the
'[international mode](#filesystem)'.

## Examples

The package contains an example of an `amigaDisk` object, which is formatted
with the old [file system](#filesystem) and a [bootable bootblock](#bootblock).
Files and directories in the root directory can easily be listed.

```{r}
list.adf.files(adf.example)
```

The disk also contains a file called 'Startup-Sequence'. This executable
script file that is run when the disk is booted from the disk. It is
comparable with the (probably better known) 'autoexec.bat' files on
IBM PC DOS systems. Let's have a little peek at this file.

```{r}
## get the file from the amigaDisk object:
startup <- get.adf.file(adf.example, "df0:s/Startup-Sequence")

## the file content is returned as raw data.
## let's convert it to text:
startup <- rawToChar(startup)

## let's show it
cat(startup)
```

As you can see this startup sequence contains mostly some comments
and will print (Echo) some text to the command line interface and
doesn't do more than that.

There is currently little support for Amiga file types in R. An
exception is the ProTracker module format, as shown in the following
example.

```{r}
## first get the file as raw data.
mod.raw <- get.adf.file(adf.example, "df0:mods/mod.intro")

## For the rest of the example we need
## the ProTrackR package
if (requireNamespace("ProTrackR", quietly = TRUE)) {
  con <- rawConnection(mod.raw, "rb")

  ## and read it as a ProTracker module
  mod <- ProTrackR::read.module(con)
  close(con)

  ## plot the first sample from the module:
  par(mar = c(5, 4, 0, 0) + 0.1)
  plot(ProTrackR::waveform(ProTrackR::PTSample(mod, 1)),
       type = "l", ylab = "Amplitude")
  
  ## and to play it, uncomment the following line:
  ## ProTrackR::playMod(mod)
}
```

The examples above show how information can be retrieved from
a virtual Amiga disk. It is also possible to put data onto the
disk. Let's start by creating a temporary directory:

```{r}
adf.example <- dir.create.adf(adf.example, "temp")
```

You can also put files from your local system into this newly
created directory. Let's try to put the &lsquo;DESCRIPTION&rsquo; file
from adfExplorer package on the virtual disk:

```{r}
adf.example <- put.adf.file(adf.example,
                            system.file("DESCRIPTION", package = "adfExplorer"),
                            "DF0:temp")
```

It is also possible to write raw data to the virtual disk in a
similar way:

```{r}
adf.example <- put.adf.file(adf.example,
                            charToRaw("This is just some text to create some content"),
                            "DF0:temp/example.txt")
```

Look here are the files we just put onto the disk

```{r}
list.adf.files(adf.example, "DF0:temp/")
```

Do you regret them? Just delete them:

```{r}
adf.example <- 
  adf.file.remove(adf.example, "DF0:temp")
list.adf.files(adf.example, "DF0:")
```
